### 2015-04-17 zoldatoff/tex/text retrieval/lecture05.tex
Лекции по курсу Text Retrieval
Неделя 2, лекции 6-8
Лекции по курсу Text Retrieval
Неделя 4. Лекции 1-4

%----------------------------------------
\subsection{Vector Space Model (VSM): Illustration}

@@

%----------------------------------------
\subsection{VSM Is a Framework}
\begin{itemize}
\item Represent a doc/query by a term vector
\begin{itemize}
\item \textbf{Term}: basic concept, e.g., word or phrase
\item Each term defines one dimension
\item N terms define an \textbf{N-dimensional space}
\item \textbf{Query vector}: $q=(x_1, \dots x_N), x_i \in \Re$ is query term weight 
\item \textbf{Doc} vector: $d=(y_1, \dots y_N), y_j \in \Re$ is doc term weight
\end{itemize}
\item $relevance(q,d) \propto similarity(q,d)=f(q,d)$
\end{itemize}

%----------------------------------------
\subsection{What VSM Doesn’t Say}
\begin{itemize}
\item How to define/select the “basic concept” – Concepts are assumed to be orthogonal
\item How to place docs and query in the space (= how to assign term weights)

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{vsm_questions.png}
\end{figure}

%----------------------------------------
\subsection{Simplest VSM = Bit-Vector + Dot-Product + BOW}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{simplest_vsm.png}
\end{figure}

@@

%----------------------------------------
\subsection{Improved VSM with Term Frequency (TF) Weighting}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{VSM_TF.png}
\end{figure}

%----------------------------------------
\subsection{IDF Weighting: Penalizing Popular Terms}
IDF — inverse document frequency
\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\linewidth]{IDF.png}
\end{figure}

%----------------------------------------
\subsection{Adding Inverse Document Frequency (IDF)}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\linewidth]{VSM_IDF.png}
\end{figure}

%----------------------------------------
\subsection{Ranking Function with TF-IDF Weighting}

@@

%----------------------------------------
\subsection{TF Transformation: BM25 Transformation}
BM = Best Matching

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{BM25_transformation.png}
\end{figure}

%----------------------------------------
\subsection{Summary}
\begin{itemize}
\item Sublinear TF Transformation is needed to
\begin{itemize}
\item capture the intuition of <<diminishing return>> from higher TF 
\item avoid dominance by one single term over all others
\end{itemize}

@@

\item Ranking function with BM25 TF ($k >= 0$):
\end{itemize}

\begin{equation*}
f(q, d) = \sum_{i=1}^N x_i y_i = \sum_{w \in q \cap d} c(w, q) \frac{(k+1) c(w, d)}{c(w, d) + k} \log \frac{M+1}{df(w)}
\end{equation*}

%----------------------------------------
\subsection{Pivoted Length Normalization}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{pivoted_length_norm.png}
\end{figure}

%----------------------------------------
\subsection{State of the Art VSM Ranking Functions}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{pivoted_length_normalization_vsm.png}
\end{figure}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{BM25_Okapi.png}
\end{figure}

>>>>

\subsection{Basic Idea}
%----------------------------------------
\subsubsection{Vector Space Model (VSM): Illustration}

@@

%----------------------------------------
\subsubsection{VSM Is a Framework}
\begin{itemize}
\item Represent a doc/query by a term vector
\begin{itemize}
\item \textbf{Term}: basic concept, e.g., word or phrase
\item Each term defines one dimension
\item N terms define an \textbf{N-dimensional space}
\item \textbf{Query vector}: $q=(x_1, \dots x_N), x_i \in \Re$ is query term weight 
\item \textbf{Doc} vector: $d=(y_1, \dots y_N), y_j \in \Re$ is doc term weight
\end{itemize}
\item $relevance(q,d) \propto similarity(q,d)=f(q,d)$
\end{itemize}

%----------------------------------------
\subsubsection{What VSM Doesn’t Say}
\begin{itemize}
\item How to define/select the “basic concept” – Concepts are assumed to be orthogonal
\item How to place docs and query in the space (= how to assign term weights)

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{vsm_questions.png}
\end{figure}

%----------------------------------------
\subsection{Simplest VSM = Bit-Vector + Dot-Product + BOW}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{simplest_vsm.png}
\end{figure}

@@

%----------------------------------------
\subsubsection{Improved VSM with Term Frequency (TF) Weighting}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{VSM_TF.png}
\end{figure}

%----------------------------------------
\subsubsection{IDF Weighting: Penalizing Popular Terms}
IDF — inverse document frequency
\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\linewidth]{IDF.png}
\end{figure}

%----------------------------------------
\subsubsection{Adding Inverse Document Frequency (IDF)}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{VSM_IDF.png}
\end{figure}

%----------------------------------------
\subsection{TF Transformation}
\subsubsection{Ranking Function with TF-IDF Weighting}

@@

%----------------------------------------
\subsubsection{TF Transformation: BM25 Transformation}
BM = Best Matching

\begin{figure}[H]
    \centering
    \includegraphics[width=0.65\linewidth]{BM25_transformation.png}
\end{figure}

%----------------------------------------
\subsubsection{Summary}
\begin{itemize}
\item Sublinear TF Transformation is needed to
\begin{itemize}
\item capture the intuition of <<diminishing return>> from higher TF 
\item avoid dominance by one single term over all others
\end{itemize}

@@

\item Ranking function with BM25 TF ($k \geqslant 0$):
\begin{equation*}
f(q, d) = \sum_{i=1}^N x_i y_i = \sum_{w \in q \cap d} c(w, q) \frac{(k+1) c(w, d)}{c(w, d) + k} \log \frac{M+1}{df(w)}
\end{equation*}
\end{itemize}

%----------------------------------------
\subsection{Doc Length Normalization}
\subsubsection{Pivoted Length Normalization}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.65\linewidth]{pivoted_length_norm.png}
\end{figure}

%----------------------------------------
\subsubsection{State of the Art VSM Ranking Functions}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{pivoted_length_normalization_vsm.png}
\end{figure}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{BM25_Okapi.png}
\end{figure}

### 2015-04-17 zoldatoff/tex/text retrieval/lecture07.tex
Лекции по курсу Text Retrieval
Неделя 4. Лекции 1-4

%----------------------------------------
\subsection{Evaluating a Set of Retrieved Docs}

@@

%----------------------------------------
\subsection{Combine Precision and Recall: F-Measure}

@@

%\pagebreak
%----------------------------------------
\subsection{Evaluating Ranking: Precision-Recall (PR) Curve}
\begin{itemize}
\item Total number of relevant documents in collection: $a+b = 10$
\item Number of retrieved documents: $a+c = 10$
\end{itemize}

@@

@@

%----------------------------------------
\subsection{How to Summarize a Ranking}
\href{http://en.wikipedia.org/wiki/Information_retrieval#Average_precision}{Average Precision} is sensitive to the rank of each relevant document:
\begin{equation*}
AveP = \dfrac{\sum\limits_{k=1}^{a+c} P(k) \cdot rel(k)}{a+b} = \sum\limits_{k=1}^{a+c} P(k) \cdot \Delta r(k)
\end{equation*}
\begin{itemize}
\item $a+c$ - number of retrieved documents
\item $a+b$ - total number of relevant documents in collection
\item $P(k)$ - the precision at cut-off $k$ in the list
\item $rel(k)$ - indicator function equaling 1 if the item at rank $k$ is a relevant document, zero otherwise
\item $\Delta r(k)$ - the change in recall that happened between cut-off $k-1$ and cut-off $k$
\end{itemize}

@@

%----------------------------------------
\subsection{Mean Average Precision (MAP)}
In case of multiple queries:
\begin{itemize}
\item MAP = arithmetic mean of average precision over a set of queries 
\begin{equation*}
MAP = \frac{\sum\limits_{q=1}^{N}AveP(q)}{N}\textit{,\; where N is the number of queries}
\end{equation*}

@@

%----------------------------------------
\subsection{Summary on Average Precision}
\begin{itemize}
\item Precision-Recall curve characterizes the overall accuracy of a ranked list
\item The \textbf{actual} utility of a ranked list depends on how many top-ranked results a user would examine
\item Average Precision is the standard measure for comparing two ranking methods
\begin{itemize}
\item Combines precision and recall
\item Sensitive to the rank of \textbf{every} relevant document
\end{itemize}
\end{itemize}

>>>>

%----------------------------------------
\subsection{Basic Measures}
\subsubsection{Evaluating a Set of Retrieved Docs}

@@

%----------------------------------------
\subsubsection{Combine Precision and Recall: F-Measure}

@@

%----------------------------------------
\newpage
\subsection{Evaluating a Ranked List}
\subsubsection{Evaluating Ranking: Precision-Recall (PR) Curve}
\begin{itemize}
\item Total number of relevant documents in collection: $a+b = 10$
\item Number of retrieved documents: $a+c = 10$
\end{itemize}

@@

\columnbreak

@@

%----------------------------------------
\subsubsection{How to Summarize a Ranking}
\href{http://en.wikipedia.org/wiki/Information_retrieval#Average_precision}{Average Precision} is sensitive to the rank of each relevant document:
\begin{equation*}
AveP = \dfrac{\sum\limits_{k=1}^{a+c} P(k) \cdot rel(k)}{a+b} = \sum\limits_{k=1}^{a+c} P(k) \cdot \Delta r(k)
\end{equation*}
\begin{itemize}
\item $a+c$ - number of retrieved documents
\item $a+b$ - total number of relevant documents in collection
\item $P(k)$ - the precision at cut-off $k$ in the list
\item $rel(k)$ - indicator function equaling 1 if the item at rank $k$ is a relevant document, zero otherwise
\item $\Delta r(k)$ - the change in recall that happened between cut-off $k-1$ and cut-off $k$
\end{itemize}

@@

%----------------------------------------
\subsubsection{Mean Average Precision (MAP)}
In case of multiple queries:
\begin{itemize}
\item MAP = arithmetic mean of average precision over a set of queries 
\begin{equation*}
MAP = \frac{\sum\limits_{q=1}^{N}AveP(q)}{N}\text{,\; where $N$ is the number of queries}
\end{equation*}

@@

%----------------------------------------
\subsubsection{Summary}
\begin{itemize}
\item Precision-Recall curve characterizes the overall accuracy of a ranked list
\item The \textbf{actual} utility of a ranked list depends on how many top-ranked results a user would examine
\item Average Precision is the standard measure for comparing two ranking methods
\begin{itemize}
\item Combines precision and recall
\item Sensitive to the rank of \textbf{every} relevant document
\end{itemize}
\end{itemize}
